Poisson approximate likelihood compared to the particle filter

### 2024-09-17 ms.Rnw

\begin{table}[ht] % The placement specifier can be [h!tbp]
\centering 
\caption{AIC for the OvOv rotavirus model, computed using two filtering methods. PAL is the Poisson approximate likelihood, implemented using the code of WWR. PF is the particle filter, implemented using the R package pomp \citep{pomppackagepaper}. Lines 1, 2 and 7 are from WWR, and the remainder are our own computations. We used $5 \times 10^4$ particles for both PF and PAL. PF was repeated 36 times to reduce the Monte Carlo variance, but this step was not necessary for PAL due to its lower Monte~Carlo variance. PF results were maximized using iterated filtering, following the approach of SBH. PAL results were maximized using coordinate gradient descent, using the code of WWR.
} 
\label{tab:ovovrealdata}
\begin{tabular}{llllr} 
  \hline
  & Method & Data & Model & AIC  \\
  \hline
  1. & PF & Rescaled counts & SBH &  \Sexpr{myround(20134.38,tab_signif)}
  \\
  2. & PAL & Counts &  WWR & \Sexpr{myround(13778.08,tab_signif)}
  \\
  3. & PAL & Counts & Re-evaluating WWR at published parameters &
    \Sexpr{myround(-2*palfilter_lik_list_wwr$log_lik + 2*7 ,tab_signif)} 
  \\
  4. & PF & Counts & Model and parameters matching line 3 &
    \Sexpr{myround(-2*logmeanexp(logLik(pfilter_wwr)) + 2*7,tab_signif)}
  \\
  5. & PF & Counts & WWR with initial values changed to SBH &
     \Sexpr{myround(-2*est_mle_vec_sbh["logLik"] + 2*7,tab_signif)}
  \\
  6. & PAL & Counts & Model and parameters matching line 5  &
     \Sexpr{myround(-2*palfilter_mif2_mle$log_lik + 2*7 ,tab_signif)}
  \\   
  7. & Benchmark & Rescaled counts  & log-ARMA$(2,1)$ & 23043
  \\
  8. & Benchmark & Counts           &  log-ARMA$(2,1)$ & \Sexpr{myround(wwr_benchmark,tab_signif)} 
  \\
 \hline
\end{tabular}
\end{table}

@@

Inspection of log-likelihood anomalies \citep{wheeler24,li24} showed that the initial conditions for the latent process in January 2001 were fixed at values which were incompatible with the trajectory of the data early in the time series \citep[Figure~3 of][]{hao24}.
By contrast, SBH fixed their initial conditions 6 years before the first measurement, giving time for the system to reach its equilibrium distribution.
Line 5 of Table~\ref{tab:ovovrealdata} incorporates the SBH specification of initial values into the model of WWR.
Comparison of lines 3 and 5 shows that this improvement enables PF to reach the values attained by PAL, and show a small improvement.
Further, at the MLE obtained by maximizing with PF, we found that PF beats PAL (comparing lines 5 and 6).

>>>>

\begin{table}[ht] % The placement specifier can be [h!tbp]
\centering 
\caption{AIC for the OvOv rotavirus model, computed using two filtering methods. PAL is the Poisson approximate likelihood, implemented using the code of WWR. PF is the particle filter, implemented using the R package pomp \citep{pomppackagepaper}. Lines 1, 2 and 7 are taken from WWR, and the remainder are our own computations. Line 3 recomputes the previously published value in Line 2, and the small difference is presumably due to rounding in Table~2 of WWR. We used $5 \times 10^4$ particles for both PF and PAL. PF was repeated 36 times to reduce the Monte Carlo variance, but this step was not necessary for PAL due to its lower Monte~Carlo variance. PF results were maximized using iterated filtering, following the approach of SBH. PAL results were maximized using coordinate gradient descent, using the code of WWR.
} 
\label{tab:ovovrealdata}
\begin{tabular}{lllllr} 
  \hline
  & Method & Data & Model & Parameters                          & AIC  \\
  \hline
  1. & PF  & Rescaled counts  & SBH   & Table~2 of SBH &  \Sexpr{myround(20134.38,tab_signif)}
  \\
  2. & PAL & Counts  & WWR   & Table~2 of WWR                      & \Sexpr{myround(13778.08,tab_signif)}
  \\
  3. & PAL & Counts  & WWR   & Table~2 of WWR  &
    \Sexpr{myround(-2*palfilter_lik_list_wwr$log_lik + 2*7 ,tab_signif)} 
  \\
  4. & PF  & Counts & WWR    & Table~2 of WWR &
    \Sexpr{myround(-2*logmeanexp(logLik(pfilter_wwr)) + 2*7,tab_signif)}
  \\
  5. & PF  & Counts & WWR modified   & Maximum likelihood &
     \Sexpr{myround(-2*est_mle_vec_sbh["logLik"] + 2*7,tab_signif)}
  \\
  6. & PAL & Counts & WWR modified   & Matching line 5 &
     \Sexpr{myround(-2*palfilter_mif2_mle$log_lik + 2*7 ,tab_signif)}
  \\   
  7. & Benchmark & Rescaled counts & log-ARMA(2,1)    &  & 23043
  \\
  8. & Benchmark & Counts & log-ARMA(2,1)      &  &
    \Sexpr{myround(wwr_benchmark,tab_signif)} 
  \\
 \hline
\end{tabular}
\end{table}

@@

Inspection of log-likelihood anomalies \citep{wheeler24,li24} showed that the initial conditions for the latent process in January 2001 were fixed at values which were incompatible with the trajectory of the data early in the time series \citep[Figure~3 of][]{hao24}.
By contrast, SBH fixed their initial conditions 6 years before the first measurement, giving time for the system to reach its equilibrium distribution.
Line 5 of Table~\ref{tab:ovovrealdata} incorporates the SBH specification of initial values into the model of WWR.
The likelihood for this modified model was then maximized using an iterated filtering procedure similar to SBH.
Comparison of lines 3 and 5 shows that this improvement enables PF to reach the AIC values attained by PAL, and show a small improvement.
Further, at the parameter values obtained by maximizing with PF, we found that PF beats PAL (comparing lines 5 and 6).
